Analyzing Fairness: Race, Recidivism, and the COMPAS Algorithm

### 2022-04-28 Paper_Draft2/Draft2.tex

\begin{longtable}[]{@{}lrrr@{}}
\caption{Incidence of recidivism by race, illustrating how a much
greater proportion (\textgreater{} 50\%) of Black defendants recidivated
than their white counterparts. \label{tab:recid table}}\tabularnewline
\toprule
Two Year Recidivism by Race & Recidivated & Survived &
Total\tabularnewline
\midrule
\endfirsthead
\toprule
Two Year Recidivism by Race & Recidivated & Survived &
Total\tabularnewline
\midrule
\endhead
African American & 1661 & 1512 & 3173\tabularnewline
Caucasian & 822 & 1278 & 2100\tabularnewline
Total & 2483 & 2790 & 5273\tabularnewline
\bottomrule
\end{longtable}

@@

Pre-processing refers to assessing the training data, and it is the most
flexible method because it has not yet trained a model that may carry
assumptions about the data. It is important to keep in mind that
pre-processing prevents assumptions in the modeling, but does not
account for the bias in data collection. With pre-processing, we use the
re-weighing algorithm from AIF360 which weights the examples in each
(group, label) combination differently to ensure fairness before
classification \citep{aif360-oct-2018}. We use a logistic regression
model for this algorithm, as it is the easiest to interpret in the given
context. After running the fairness metrics using the pre-processing
algorithm, we are able to compare our results to the baseline metrics
from the previous section.

Another approach, in-processing, explores how bias can be targeted while
building a model. It assesses how we build learning algorithms, which
often prioritize accuracy over fairness. The technique we use is the
Meta Classification algorithm, which accounts for the fairness metric as
part of the input and returns a classifier optimized by that particular
metric. Similarly to pre-processing, we compare the results of our
in-processing methods with both the baseline and the pre-processing to
gauge which method so far has better individual or group fairness.

Our last approach, post-processing, makes transformations on model
outputs. Similarly to the pre-processing, it is flexible because it does
not work directly with the algorithm, but instead has direct access to
all the predicted classification values. Because the algorithms are
often so complex and difficult to interpret, post-processors are trained
in a black box. To get around this, we use the PostProcessingMeta class
from AIF360 which combines the training and prediction for a random
estimator and the post-processor while simultaneously splitting the data
set \citep{aif360-oct-2018}. Based on these results, we look
comparatively at all the approaches to determine which we assess to be
the most fair.

>>>>

\begin{longtable}[]{@{}lrrr@{}}
\caption{Incidence of recidivism by race, illustrating how a much
greater proportion (\textgreater{} 50\%) of Black defendants recidivated
than their white counterparts. \label{tab:recid table}}\tabularnewline
\toprule
Two Year Recidivism by Race & Recidivated & Survived & Total \\
\midrule
\endfirsthead
\toprule
Two Year Recidivism by Race & Recidivated & Survived & Total \\
\midrule
\endhead
African American & 1661 & 1512 & 3173 \\
Caucasian & 822 & 1278 & 2100 \\
Total & 2483 & 2790 & 5273 \\
\bottomrule
\end{longtable}

@@

Pre-processing refers to mitigating bias within the training data, and
it is the most flexible method because it has not yet trained a model
that may carry assumptions about the data. It is important to keep in
mind that pre-processing prevents assumptions in the modeling, but does
not account for the bias in data collection. Training data is where bias
is most likely to be introduced. We use the reweighing pre-processing
algorithm from AIF360 which assigns weights to the data. ``The advantage
of this approach is, instead of modifying the labels, it assigns
different weights to the examples based upon their categories of
protected attribute and outcome such that bias is removed from the
training dataset. The weights are based on frequency counts. However as
this technique is designed to work only with classifiers that can handle
row-level weights, this may limit your modeling options''
\citep{Ronaghan2019AI}. We used a logistic regression model for this
algorithm, as it is the easiest to interpret in the given context. After
running the fairness metrics using the pre-processing algorithm, we were
able to compare our results to the baseline metrics from the previous
section.

In-processing mitigates bias in classifiers while building a model. A
classifier ``is an algorithm that automatically orders or categorizes
data into one or more sets'' \citep{baxter2021what}. The in-processing
technique we use is the prejudice remover algorithm, which accounts for
the fairness metric as part of the input and returns a classifier
optimized by that particular metric. In order to do this, we first
needed to convert our data frame into a data type called a
BinaryLabelDataset. Similarly to pre-processing, we compared the results
of our in-processing methods with both the baseline and the
pre-processing to gauge which method so far has better individual or
group fairness.

The prejudice remover is a method for reducing indirect prejudice
(i.e.~how COMPAS is racially biased because it uses proxy variables for
race). The prejudice remover implements two different regularizers, one
to avoid overfitting and one to enforce fair classification. The
prejudice remover regularizer works by minimizing the prejudice index, a
mathematical equation for quantifying fairness defined by Kamishima et.
al.~This in turn enforces a classifier's independence from sensitive
information (e.g., race).

Our last approach, post-processing bias mitigation, is used after
training a model. Post-processing algorithms equalize the outcomes
(i.e., predicted classification values) to mitigate bias instead of
adjusting the classifier or the training data \citep{baxter2021AI}. We
use calibrated equalized odds, which ``optimizes over calibrated
classifier score outputs to find probabilities with which to change
output labels with an equalized odds objective'' {[}insert aif360
website citation here{]}. An equalized odds objective constrains
classification algorithms such that no error type (false-positive or
false-negative) disproportionately affects any population subgroup; both
groups, in our case both white and Black defendants, should have the
same false-positive and false-negative rates. Through the calibrated
equalized odds method, we want to decrease bias while also maintaining
calibration \citep{pleiss2017fairness}.
\href{https://medium.com/analytics-vidhya/calibration-in-machine-learning-e7972ac93555}{Calibration}
refers to improving a model so that the distribution of predicted
outcomes is similar to the distribution of observed probability in the
training data.

### 2022-04-28 final_metrics_analysis.Rmd

```{python pre-processing accuracy score} 
acc_REW = accuracy_score(y_test, y_pred_REW)
print(f'[Reweighting] The test accuracy of the algorithm is: {acc_REW: .2%}')
```

```{python pre-processing confusion matrix}
cf_matrix = confusion_matrix(y_test, y_pred_REW)
make_confusion_matrix(cf_matrix, "[Reweighting]")
plt.show()
```

The highest percentage in this confusion matrix is for the true negative rate, defendants who were predicted to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model. This means that the re-weighing model marginally improved the baseline model accuracy in predicting people who did not recidivate. However, the true positive rate, or those who were predicted to recidivate and did recidivate, is 24.72%, 1.13% lower than that of the baseline model. This means that the re-weighing model slightly lowered the accuracy of the baseline model in predicting people who did recidivate. As a result, the overall reweighting model accuracy remains about the same as the baseline model accuracy. The false negative and false positive rates, or the predictions that proved to be incorrect, are 23.20% and 15.85%, respectively. The re-weighing model increased the number of false negatives by 1.14% and decreased the number of false positives by 0.83%. We are most concerned about the false positives, which indicate defendants who are predicted to recidivate and do not actually recidivate. While there is a slight improvement in the percent of false positives, it is still relatively high at 15.85%, indicating that we should further precede with additional bias mitigation techniques.

@@

We can see clear effects of the reweighting approach in the recall scores, which are now all very similar around 0.5. This is a marked improvement from the baseline model, where the model only correctly identified white defendants who recidivated 20% of the time. The reweighting model now correctly identifies defendants of any race who recidivate about half of the time. Notably, the reweighting also lowered the recall score for Black defendants.

Whereas with the baseline model all three precision scores were relatively similar, with the reweighting model we see the precision score for Black defendants is the highest. The models predictions for Black defendants who recidivate are correct about 66% of the time. In contrast, these predictions are correct for white defendents only 50% of the time, and correct for all defendants 60% of the time. The increase in precision for Black defendants shows the effects of the bias mitigation, as fewer Black people are being incorrectly predicted to recidivate.

@@

print(f'[Reweighting] The average odds difference is {avg_odds_diff_RW: .2}')
```

@@

print(f'[Reweighting] The disparate impact ratio is {disp_impact_ratio_RW: .2}')
```

The reweighting approach also improves the disparate impact ratio, bringing the value almost to 1, which conveys absolute fairness.

### Plot Fairness Metrics (Reweighting)

```{python pre-processing fairness metrics plot}
plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, avg_odds_diff_RW, disp_impact_ratio_RW], 'Reweighting')
plt.show()
```

The orange bars in these graphs represent the new values of the fairness metrics after implementing the reweighting approach. Like we saw above, reweighting the data improved the fairness in all four of these metrics. All four values are now within the range of values considered fair. Statistical parity difference and disparate impact ratio still imply a slight benefit for white defendants, while equal opportunity difference and average odds difference now imply a slight benefit for Black defendants.

@@

After the prejudice remover, the accuracy of the model is fairly similar across the board of races, but the model's accuracy is only 38.21%, which is much lower than the baseline and reweighting models. This accuracy score means that the model makes accurate predictions only 38% of the time.

>>>>

```{python pre-processing accuracy score} 
acc_REW = accuracy_score(y_test, y_pred_REW)
print(f'[Reweighing] The test accuracy of the algorithm is: {acc_REW: .2%}')
```

```{python pre-processing confusion matrix}
cf_matrix = confusion_matrix(y_test, y_pred_REW)
make_confusion_matrix(cf_matrix, "[Reweighing]")
plt.show()
```

The highest percentage in this confusion matrix is for the true negative rate, defendants who were predicted to not recidivate and actually did not, at 36.24%, 0.83% higher than the true negatives in the baseline model. This means that the re-weighing model marginally improved the baseline model accuracy in predicting people who did not recidivate. However, the true positive rate, or those who were predicted to recidivate and did recidivate, is 24.72%, 1.13% lower than that of the baseline model. This means that the re-weighing model slightly lowered the accuracy of the baseline model in predicting people who did recidivate. As a result, the overall reweighing model accuracy remains about the same as the baseline model accuracy. The false negative and false positive rates, or the predictions that proved to be incorrect, are 23.20% and 15.85%, respectively. The re-weighing model increased the number of false negatives by 1.14% and decreased the number of false positives by 0.83%. We are most concerned about the false positives, which indicate defendants who are predicted to recidivate and do not actually recidivate. While there is a slight improvement in the percent of false positives, it is still relatively high at 15.85%, indicating that we should further precede with additional bias mitigation techniques.

@@

We can see clear effects of the reweighing approach in the recall scores, which are now all very similar around 0.5. This is a marked improvement from the baseline model, where the model only correctly identified white defendants who recidivated 20% of the time. The reweighing model now correctly identifies defendants of any race who recidivate about half of the time. Notably, the reweighing also lowered the recall score for Black defendants.

Whereas with the baseline model all three precision scores were relatively similar, with the reweighing model we see the precision score for Black defendants is the highest. The models predictions for Black defendants who recidivate are correct about 66% of the time. In contrast, these predictions are correct for white defendents only 50% of the time, and correct for all defendants 60% of the time. The increase in precision for Black defendants shows the effects of the bias mitigation, as fewer Black people are being incorrectly predicted to recidivate.

@@

print(f'[Reweighing] The average odds difference is {avg_odds_diff_RW: .2}')
```

@@

print(f'[Reweighing] The disparate impact ratio is {disp_impact_ratio_RW: .2}')
```

The reweighing approach also improves the disparate impact ratio, bringing the value almost to 1, which conveys absolute fairness.

### Plot Fairness Metrics (Reweighing)

```{python pre-processing fairness metrics plot}
plot_fair_metrics([stat_par_diff_RW, eq_opp_diff_RW, avg_odds_diff_RW, disp_impact_ratio_RW], 'Reweighing')
plt.show()
```

The orange bars in these graphs represent the new values of the fairness metrics after implementing the reweighing approach. Like we saw above, reweighing the data improved the fairness in all four of these metrics. All four values are now within the range of values considered fair. Statistical parity difference and disparate impact ratio still imply a slight benefit for white defendants, while equal opportunity difference and average odds difference now imply a slight benefit for Black defendants.

@@

After the prejudice remover, the accuracy of the model is fairly similar across the board of races, but the model's accuracy is only 38.21%, which is much lower than the baseline and reweighing models. This accuracy score means that the model makes accurate predictions only 38% of the time.
